```{r run-all-Melbourne-code}
area_name <- "Melbourne"
lga_level_data <- run_for_area(area_name)

```
# `r area_name`

```{r}
htmltools::withTags(table(
  id = "lga-summary",
    tr(
      td(id = "lga-2", sprintf("Housing target:", lga_level_data$profitable_units)),
      td(id = "lga-1", paste0(lga_level_data$mm_target, " homes/year"))
    ),
    tr(
      td(id = "lga-2", "Total housing capacity:"),
      td(id = "lga-1", lga_level_data$mm_zoned_capacity)
    ),
  tr(
      td(id = "lga-2", "Land to build new housing:"),
      td(id = "lga-1", paste0(lga_level_data$mcgs_of_space, " MCGs"))
    ),
  )
)

```

```{r}
lga_level_data$mm_zoning_map
```


## More homes in `r lga_level_data$area_name`  

Melbourne is unique amongst inner and middle-LGAs in that it is delivering a singificant supply of homes to meet the city's housing needs. But Melbourne cannot simply rely on Brownfield redevelopment to meet its targets. The best located land in Melbourne is in areas where it is currently illegal to build. Suburbs such as East Melbourne, Carlton, Parkville and Kensington are close to parks, public transport and schools, and yet Melbourne instead relegates density to some of the most polluted and unpleasant areas of the LGA. 

Through broad transit-oriented upzoning and development, `r area_name` can deliver housing abundance not just on traffic sewers like Fisherman's Bend or City Road, but also near Fitzroy Gardens and Royal Park.

In `r area_name` missing middle upzoning will have the following effect:

```{r}
options(scipen=999)

total_lots_upzoned <- prettyNum(round(parse_number(lga_level_data$mm_number) + parse_number(lga_level_data$rgz_number) + parse_number(lga_level_data$grz_number)), big.mark = ",")

```

| Height      	|  Number of lots upzoned	|
|:-------------------------	|---:	|
| 6 storeys 	|  `r lga_level_data$mm_number`  	|
| 4 storeys  	|  `r lga_level_data$rgz_number` 	|
| 3 storeys  	|  `r lga_level_data$grz_number`	|
| **Total** | **`r total_lots_upzoned`**  	|


## Housing targets

Even though there are __`r lga_level_data$mm_zoned_capacity` total possible units__ that could be built in `r lga_level_data$area_name`, it's impossible for them to be built all at once. It is also not reasonable to assume that all possible units will ever be built, which is [why excess zoned capacity is important](#). 

For each lot in `r area_name`, we calculated whether it would be profitable to build housing under upzoning. We estimate that there are __`r lga_level_data$profitable_units` viable units__ that could be built in `r area_name`. 

Based off our data, in order to contribute its fair share to the the statewide target of 80,000 new homes per year, `r area_name` must build  **`r lga_level_data$mm_target` units** per year.^1^ This target is in addition to any housing delivered as part of the Arden and Fisherman's bend. 

City of Melbourne is likely to meet or exceed these targets, but these levels are seen as the bare miimum to be delivered as part of any housing strategy. 

## Planning barriers

While there has been significant housing construction ub Melbourne, restrictive zoning still holds back many areas. Here is the current zoning map for `r lga_level_data$area_name`:

```{r}
lga_level_data$existing_zoning_map

```


Increasing zoned capacity in `r lga_level_data$area_name` by `r lga_level_data$zoned_capacity_x`, with a focus on density around transport hubs, will enable more homes to be built where people want to live.

### Heritage should not be a barrier to housing abundance

Of the `r lga_level_data$profitable_units` potentially profitable apartments that could be built in `r lga_level_data$area_name`, `r lga_level_data$heritage_no` of these potential future homes are on lots subject to heritage controls. 

If heritage and neighbourhood character protections in `r lga_level_data$area_name` are too far-reaching, it could potentially affect `r lga_level_data$area_name`'s ability to meet its housing targets. 

We have outlined potential remedies for this [earlier in the report](#report-link).

**An image here should end the article.**

[Return to LGA list](#report-link)