# Inner-city housing targets

More housing should be built in areas where people want to live, and where greater density can be supported. 

As such, we have allocated half of Victoria's [annual target of 80,000 new dwellings](https://content.vic.gov.au/sites/default/files/2023-09/DTP0424_Housing_Statement_v6_FA_WEB.pdf) to 19 amenity- and infrastructure-rich inner-city LGAs where demand for apartment and townhouse living is high, and where supply by the private market will be most viable. 

These targets exist in addition to housing built through greenfield development and precincts such as Fisherman's Bend and Suburban Rail Loop stations. 

We have calculated housing targets for inner and middle Melbourne. 

Targets are highest in places like Boroondara where density is low due high numbers of single-family homes on large lots near public transport, where there is strong demand for apartments. 

Targets are lower in areas such as Maribyrnong, where apartments sell for much less and missing middle housing is harder to build. 

There are also some areas like Port Phillip, where demand is high and yet there simply isn't enough suitable land to build missing middle housing. 

More housing could potentially be provided in these LGAs through rezoning of well-located industrial and commercial areas. However, for the purposes of this report, we are only considering land on which housing is currently permitted.


```{r,fig.height=7,fig.width = 8}

map_data <- strayr::read_absmap("lga2022") %>% 
  filter(state_name_2021 == "Victoria") %>% 
    mutate(lga_name_2022 = str_remove_all(lga_name_2022, "\\s*\\(.*?\\)\\s*")) %>% 
    mutate(lga_name_2022 = if_else(lga_name_2022 == "Moreland","Merri-bek",lga_name_2022)) %>%  # Moreland changed name in 2021
  inner_join(lga_zoning_numbers) 

#devtools::install_github("yutannihilation/ggsflabel")
map_data%>% 
  mutate(target = paste0(lga_name_2022,"\n",
                         prettyNum(round(mm_target,-2),
                                   big.mark = ","),
                         " homes")
         ) %>% 
  ggplot()+
  geom_sf(colour = yimby_colours$green_palette[4])+
  ggthemes::theme_map()+
  ggsflabel::geom_sf_label_repel(aes(label = target ),
                                 size = 2,
                                 fill = yimby_colours$background)+
    labs(title = "Missing Middle Melbourne yearly housing targets",
         x = element_blank(),
         y = element_blank()) +
  theme_yimby_mel_caption(text_size = "small",plot_type = "map")+
  theme(plot.margin = margin(2, 15,2, 15))+
  theme(panel.background = element_rect(fill = "transparent", colour = NA),  
        plot.background = element_rect(fill = "transparent", colour = NA))


```

## Why Melbourne needs targets

With the exception of City of Melbourne, inner- and middle–ring councils are not doing enough to house Melbourne's future population. This has contributed to the city's high reliance on greenfield development, pushing more and more Melburnians onto our urban fringe. 


```{r fig.height = 8}

pop_change_by_lga <-lga_pop_change_importer(lgas) %>% 
  filter(year == 2022)

pop_change_by_lga %>% 
  mutate(type = if_else(lga_name_2021 == "All of Melbourne","y","n"),
         lga_name_2021 = if_else(lga_name_2021 == "Melbourne","City of Melbourne",lga_name_2021)) %>% 
  ungroup() %>% 
  mutate(change = change - 1, 
         lga_name_2021 = fct_reorder(lga_name_2021,change)) %>% 
  ggplot(aes(x = lga_name_2021, 
             y = change,
             fill = type))+
  geom_bar(stat = "identity")+
  coord_flip()+
  theme_yimby_mel_caption("Source: ABS regional populations",text_size = "small")+
  labs(title = "Middle and inner-suburban suburbs are not keeping up with\nMelbourne's need for more housing",
       subtitle = "Population change 2014-2022 (%)",
       y = element_blank(),
       x = element_blank())+
  scale_y_continuous(labels = percent_format(accuracy = 1))+
  theme(legend.position = "none")+
  geom_hline(yintercept = 0)
```

Iconic cities across the world such as Paris have a predominant built form that is medium-rise and close to public transport. Each building provides a large number of dwellings thanks to an absence of setback requirements.

The land use patterns that our city needs—ground floor commercial and retail spaces, internal courtyards, and large public enclosed squares and parks—are largely absent, because it is broadly banned from being built.

The Missing Middle reforms would enable this sort of development across our entire city, for the benefit of all Melburnians: current and future.

## Calculating targets

There is no use having a housing target if there is no land to build on, or if there is nobody who wants to live in these new homes. 

It's common practice for planners to calculate "zoned capacity" which is the theoretical number of homes that could be built across an LGA if every lot were developed to maximum permitted density. 

The problem with zoned capacity is that it is based on unrealistic assumptions and timelines. In many cases, it may be technically possible to increase density on a given lot, but not financially viable due to a high cost of land and limited yield due to restrictive zoning.

To account for this, YIMBY Melbourne's model accounts not just for an increased zoned capacity, but also for the viability of development, accounting for demand, prices, and development costs. 

To do this, we have used a precise seven-stage methodology: 

1) Sites that are parks, schools, hospitals, or already have apartments are excluded.
2) Any lot zoned "Missing Middle" is considered suitable for 6 storeys of housing with high lot coverage. The potential yield of these properties is calcualted based on the yield achieved by 6 storey Nightingdale developments in Melbourne.
3) Potential yield for RGZ, GRZ, and NRZ lots is based on the density calculated by [Professor Michael Buxton](https://cur.org.au/project/melbourne-8-million-matching-land-supply-dwelling-demand/).
4) For each lot, we calculated the cost of demolishing the existing home plus the cost of building 85 square meter GFA apartments in its place. Taller buildings are more expensive to construct owing to increased structural requirements. Because the Residential Growth Zone has parking requirements, the cost of underground car parking was added to construction costs for lots in this zone. The full cost calculation formula is available [here](https://github.com/yimbymelbourne/missing_middle_2/blob/main/R/functions/find%20profitable_apartments_for_zoning_targets.R). 
5) The value of existing properties was calculated using a regression of past house sale prices against features of the land such as the current zone, lot size, distance from CBD, SA3, and distance from public transport. The full regression terms are available [here](https://github.com/yimbymelbourne/missing_middle_2/blob/main/R/R_regressions/03run_regression.R). 
6) The likely price of apartments sold in a development on a given lot was calculated using a similar regression as above. 
7) For each site, the profitability of a development was calculated at various heights within Missing Middle zoning. Each site where profit is greater than 0 contributes to that LGA's housing target. 

Our model does not expect development to occur on every viable site. But councils with more viable sites are allocated a higher share of the state's total housing target.  

Existing heritage, neighbourhood character, and design and development overlays in place over large swathes of our inner-suburbs should not be an excuse for councils to miss their housing targets. Councils with far-reaching overlays will need to consider how best to achieve their housing targets against the tradeoffs associated with maintaining high opportunity-cost overlays.

If councils are unable to meet housing targets within current parameters, they should be given the ability to re-evaluate their overlays.

In the case where high-value heritage maintenance is deemed more important than abundant medium-density housing, councils should be able to apply for higher density on lots zoned MMZ in order to meet their targets.

Melbourne's Missing Middle Housing Targets should be updated by government every year, and if apartment prices fall relative to other areas, then the housing target will automatically fall as it will become clear that development is no longer profitable in that area.

The best way for a LGA to reduce its target is to make housing cheaper. And the best way to make housing cheaper is to make it abundant. 

This system of enforced housing targets is designed to create a virtuous cycle that drives down prices across the entire city, creating a more affordable and equitable Melbourne for all.

```{r}
tbl_2_body <- lga_zoning_numbers %>% 
  select(` ` = lga_name_2022,
         `New zoned capacity where development is viable` = profitable_new_dwellings_net,
         `Yearly new home target` = mm_target,
         `Local target as a share of city-wide target` = share_of_mm_profitable_apartments,
         `Local target relative to other LGAs (scaled to existing population)` = target_relative_to_other_missing_middle_lgas
         ) %>% 
  arrange(desc(`Local target relative to other LGAs (scaled to existing population)`))

tbl_2_total <- tbl_2_body %>% 
  summarise(across(c(`New zoned capacity where development is viable`,
                            `Yearly new home target`
                     ),sum)) %>% 
  mutate(` ` = "Total")
  
tbl_2_body%>%
  bind_rows(tbl_2_total) %>% 
  gt() %>% 
  fmt_number(c(`New zoned capacity where development is viable`,
               `Yearly new home target`),n_sigfig = 3) %>% 
  fmt_percent(columns = c(`Local target as a share of city-wide target`),decimals = 1) %>%
  fmt_percent(columns = c(`Local target relative to other LGAs (scaled to existing population)`),decimals = 0) %>% 
tab_options(
    table.background.color = yimby_colours$background# Change "lightblue" to any color you prefer
  ) %>% 
    tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = everything() %>% tail(1)
    )
  ) %>%
    fmt_missing(
    columns = everything(),
    missing_text = "-"
  )
```


## Enforcing targets

There's currently no mechanism to ensure that councils actually build the housing for which they claim to have zoned capacity. The Victorian Government should follow the advice of [Infrastructure Victoria](*) and the [Productivity Commission](*) recommendations to enforce housing targets through  attach carrots and sticks to housing targets. 

**NOTE: where are these recommendations found? Add links.**

In order for our state's planning and housing reform to be successful, it is necessary to create a robust, outcomes-oriented structure for housing target enforcement. To deliver a high number of well-located homes across Melbourne requires more than just non-binding capacity targets that look good on paper. Rather, we need binding material construction targets that look good in real life. People do not live inside housing capacity; they live inside houses. And Councils must be kept accountable to that fact.


### Carrots

LGAs who meet or exceed their housing targets should be given financial rewards by the State Government.

In order to generate the revenue required for these rewards, the state should levy a Windfall Gains Tax upon the first sale of any upzoned residential property after upzoning has taken place. This tax is outlined in YIMBY Melbourne's 2023 report, [*Melbourne's Missing Middle*](https://yimby.melbourne/missing-middle), and is based on an conservative assumption that upzoning will raise a given property's value by 4%, thanks to the increased capacity for development. We propose taxing 30% of this 4% improvement.

**MMM WGT Chart**

For a house worth \$1m, a 4% improvement is \$40,000. \$12,000 of that uplift should be paid in tax, with the owner keeping the remaining \$28,000 windfall gain. Our model proposes the upzoning of **X** properties total, with the WGT to be applied at first sale, regardless of the timing of the transfer. With the median value of houses in the area covered by this report at Y, we estimate a revenue of \$**Z** billion over the lifespan of Plan Melbourne 2050. 

A minimum of 70% of this revenue should be hypothecated to social housing builds. A portion of remaining revenue should be used to reward councils who exceed their housing targets, paid out annually and used by those councils to improve infrastructure and services for the people who live there. 

This reward could be additionally supplemented through State Government distribution of the Commonwealth's New Home Bonus to the councils in which homes are built in excess of their respective targets. 

Furthermore, the council rate cap should be abolished to empower councils to make more proactice decisions regarding expanding local infrastructure and services.


### Sticks

Councils that fail to meet their housing targets three years in a row should receive a financial penalty. 

This could take several forms. We outline three basic kinds of incentives below:

1. Councils that fail to meet their targets should be required to pay an annual 'Housing Remedy' to the state government. The proceeds of this should be hypothecated toward Homes Victoria (the state government housing agency). To increase political salience of the 'Housing Remedy', we suggest it take the form of a line-item added to every ratepayer's annual bill. 
2. Councils that fail to meet their targets should be made ineligible for select tranches of state government infrastructure grants until they course-correct. 
3. Suspend councillor call-in powers. All planning applications will be delegated to council staff.

Targets should be tracked consistently through a combination of development and building approvals, commencements and completions. In the case that a council believes its target was not met due to external factors, they should be able to appeal the penalty. 